---
title: "BGGN213-lab4"
author: "Lisanne Stouthart"
date: "2024-10-08"
output: html_document
---

## Lecture
```{r}
x <- 1:5
x + 100
y <- c(100,1)
y
x + y
```

## Start
```{r}
source("http://thegrantlab.org/misc/cdc.R")

cdc
view(cdc)

head(cdc$height)
tail(cdc$height)
```

## Q1
How would you “argue” with the tail() function to print out the last 20 weight
values? Provide your code below:
```{r}
last_20_weights <- tail(cdc$weight, 20)
print(last_20_weights)
```

## Q2
Make a scatterplot of height vs weight using the plot() function. Add the code you
used to generate this plot here:
```{r}
plot(cdc$weight, cdc$height,
     main = "Scatterplot of Height vs Weight",
     xlab = "Height (inches)",
     ylab = "Weight (pounds)",
     pch = 1,
     col = "black")
```

## Q3 + Q4
Do height and weight appear to be associated? If so are they positively associated
or negatively associated? What is the Pearson correlation value for height and weight?

A correlation close to +1 indicates a strong positive association.
A correlation close to -1 indicates a strong negative association.
A correlation around 0 suggests little to no linear association.

A correlation coefficient of 0.1 is thought to represent a weak or small association; 
a correlation coefficient of 0.3 is considered a moderate correlation; 
and a correlation coefficient of 0.5 or larger is thought to represent a strong or large correlation.
```{r}
# Calculate the correlation coefficient
correlation <- cor(cdc$weight, cdc$height, use = "complete.obs")
print(correlation)
# value = 0.5553222 -> so little to no linear association.
```

## Interim
```{r}
hist(cdc$weight)
hist(cdc$height)
height_m <- cdc$height * 0.0254
```

## Q5
Create a new object weight_kg that records weight in kilograms. Provide your code
below:
```{r}
weight_kg <- cdc$weight * .454
head(weight_kg)
```

## Q6
Create a new object bmi and make a plot of height vs BMI. Provide your code below
and comment on whether height and BMI seem to be associated?
```{r}
bmi <- (weight_kg) / (height_m^2)

plot(cdc$height, bmi,
     main = "Scatterplot of Height vs BMI",
     xlab = "Height (inches)",
     ylab = "BMI",
     pch = 1,  
     col = "black") 
```

## Q7
Are height and BMI strongly associated? What are their correlation value?
```{r}
correlation <- cor(cdc$height, bmi, use = "complete.obs")
print(correlation)
#Correlation = 0.03251694 -> so no correlation
```

## Q8
Can you use this summing of a logical vector approach to find out how many obese
individuals there are in the dataset? Provide your code and answer below:
```{r}
head(bmi >= 30, 100)
eg <- c(TRUE, TRUE, FALSE, FALSE)
sum(eg)

num_obese <- sum(bmi >= 30)
print(num_obese) #answer = 3897

sum(bmi >= 30)/length(bmi)
(sum(bmi >= 30)/length(bmi)) * 100
round( (sum(bmi >= 30)/length(bmi)) * 100, 1)
```

## Q9
Use bracket notation to make a scatterplot of height and weight for the first 100
respondents. Provide your code below.
```{r}
cdc[567, 6] #answer = 160
cdc[1:10, 6]
cdc[1:10, ]

plot(cdc$weight[1:100], cdc$height[1:100],
     main = "Scatterplot of Height vs Weight (First 100 Respondents)",
     xlab = "Weight (lbs)",
     ylab = "Height (inches)",
     pch = 1,  
     col = "black") 

#all working
#cdc[1:100, ]$height
#cdc[1:100, "height"]
#cdc[1:100, 5]

```

## Q10
How many obese individuals are male in the full dataset? Give your code and
answer below:
```{r}
obese_logical <- bmi >= 30
unique(cdc$gender)
num_obese_males <- sum(obese_logical & (cdc$gender == "m"))
print(num_obese_males) #n = 1961

```




